Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-07-31 surfaces-write-up/surface-write-up.tex
responses/fixes to todos

Aaron: In general, I like the idea of having proofs self contained,
so that one should only refer to the proof statement, not the
body of the proof.
I'm not very strongly tied to this idea though, and I think it's fine. I think it's fine.
}
We note that $\Sigma$ has extremal rays given by the lattice points 
\begin{equation}\label{defn:e-i-proj}
	e_i = \left(\ell_i \bida_i, - \alpha_1 \ell_i \bida_i, \ldots
-\alpha_{i-1} \ell_i \bida_i, \ell_i \sum_{j\ne i} \alpha_j \bida_j,
-\alpha_{i+1} \ell_i \bida_i, \ldots, -\alpha_n, \ell_i \bida_i \right)
\end{equation}
for each $i\in \{1, \ldots n\}$.
Therefore, applying Proposition ~\ref{prop:cone-generation}, we see $R_D$ is generated in degrees less than
\[
	\sum_{i=1}^n \ell_i a_i.
\]
\end{proof}

@@

From this we can deduce that a set of extremal rays of $\Sigma$ are given as follows:
For $i \in \Te(D)$, define $e_i \in R_D$ in degree $d_i = \ell_i \gcd(a_i, b_i)$ by 
\begin{equation}\label{eqn:epsilon_i}
	e_i := (-\alpha_1 d_i, \ldots, - \alpha_{i-1} d_i, \frac{d_i}{\bida_i + \bidb_i}\sum_{k \ne i}
	\alpha_k (\bida_k + \bidb_k), - \alpha_{i+1} d_i, \ldots, - \alpha_n d_i)
\end{equation}

\noindent
For $i \in \Tp(D)$ and $j \in \Tm(D)$, with $i < j$, define $e_{i,j}\in R_D$ in degree $d_{i,j} = \ell_{i, j}(\bida_i \bidb_j - \bida_j \bidb_i)$ by
\begin{equation}\label{eqn:epsilon_i,j}
	\begin{split}& e_{i,j} := (-\alpha_1 d_{i,j}, \ldots -\alpha_{i-1} d_{i,j}, \bidb_j \sum_{k \ne i,j} \alpha_k \bida_k - \bida_j \sum_{k \ne i, j}
	\alpha_k \bidb_k, -\alpha_{i+1} d_{i,j}, \\
	&\ldots -\alpha_{j-1} d_{i,j}, -\bidb_i \sum_{k \ne i, j} \alpha_k \bida_k + \bida_i \sum_{k \ne i, j}
	\alpha_k \bidb_k, -\alpha_{j+1}d_{i,j}, \ldots, - \alpha_{n}d_{i,j}).
	\end{split}
\end{equation}
Note that the above expression makes sense when $i < j$, but one
can similarly define $e_{i,j}$ when $j < i$.  
\todo{Peter: I changed the above expression,
but I'm still feel a little dubious that this is easier to read...

Aaron: I agree this is harder to read. Why don't we just write
$e_{i,j} = a e_i + b e_j$ (where of course you replace a and b by explicit
constants). Wouldn't that be much cleaner? Also, I think we should
explicitly state what the small rearrangement is to switch 
$i$ and $j$, but perhaps it won't be relevant if you write $e_{ij} $
as a linear combination of $e_i, e_j$.}

@@

To find such a $\beta_i$,
\todo{Aaron: Shouldn't we say something above like, the kernel
	is bounded by the maximum of something over all $i$,
	so for the remainder of the proof, we fix an $i$.
	Also, I think it would help if things
	that depend on $i$, such as $E,\sigma,\lambda$
have an $i$ subscript.}
we simply write $f_i$ as a 
polynomial in $f_1, \ldots f_4$. Furthermore, these
generate all relations, since they allow us to reduce any $\prod_{j =
1}^n f_j^{c_j}$ to the canonical form where $r_i = - \lfloor d \alpha_i
\rfloor$ whenever $i > 4$.

@@

Let $\omega\in \Omega'$ correspond to the monomial $y_i \prod_{j=1}^n
{y_j}^{c_j}$.  Then we can write $\omega$ as a sum of $r_j e_j$'s for $j\in
\Te$ and $r_{j,k}e_{j,k}$ for $j\in \Tp, k\in \Tm$.  
For convenience, let $e_j$ (when it exists) be in degree $d_j$ with $j^{th}$
component $-\alpha_j d_j + \kappa_j$ for some $\kappa_j \in \bq$ and let $e_{j,k}$ (when it exists) lie in
degrees $d_{j,k}$ with $j^{th}$ component $-\alpha_j d_{j,k} + \kappa_{j,k}'$ and $k^{th}$
component $-\alpha_j d_{j,k} + \kappa_{j,k}''$ for $\kappa_{j,k}', \kappa_{j,k}'' \in \bq$
By a careful computation, we can find $\lambda \in \Sigma$ 
\todo{Aaron: Is it at all possible to write down what $\lambda$ and the $r_{i,j}$ are?
I suspect we should be able to write them in a couple lines, which would
definitely clarify how we're getting the bound on degree $E$. While I agree
it's fine to hide the computation of the $r_{i,j}$ it would still clarify matters
to say what the $r_{i,j}$ and $\lambda$ are.
\\
Peter: There is not a unique choice of the $r_{i,j}$'s.  When we write out $\sigma$ in terms of the $e_i$'s and $e_{i,j}'s$ with coefficients $C_i$ and $C_{i,j}$, any choice of $r_{i,j}$'s that are at most equal to the coefficient $C_{i,j}$ and such that $\sum_{j\in T_-} r_{i,j}\kappa'_{i,j}=1$ will work.

Aaron: OK. In that case, why don't you say that? Since I didn't understand those constraints on $r_{i,j}$, I think
the reader would probably also be lost without explicitly writing those constraints.
Also, what about $\lambda$? Is that uniquely determined?}
and $r_{i,j} \in \bq$ (in general such a choice may be very non-unique) so that we can write
\begin{equation}\label{eqn:sigma-in-terms-of-lambda/E}
	\sigma = \lambda + E
 \end{equation}
with $E$ of the form
\begin{equation}\label{eqn:hirz-E-translation}
	E = \begin{cases}
	\frac{1}{\kappa_i} e_i  	&  \mbox{ if }i\in \Te\\
	\sum_{j\in \Tm} r_{i,j} e_{i,j}: r_{i,j}\in \bq, \sum_{j\in \Tm} r_{i,j}\kappa_{i,j}' = 1 & \mbox{ if } i\in \Tp\\
	\sum_{j\in \Tp} r_{j,i} e_{j,i}: r_{j,i} \in \bq, \sum_{j\in \Tm} r_{j,i} \kappa_{i,j}'' = 1 	& \mbox{ if } i\in \Tm\\
	\end{cases}.
\end{equation}

\noindent
From this, we can further deduce that for such an $E$,
\[
	\deg(E) \le \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm \end{cases}.
\]
Let $M$ be the set of monomials terms of $\beta_i$.  Then for $\mu = \prod_{j=1}^n
{f_i}^{c_i}\in M$, we
\todo{Aaron: What is $M$? \\
Peter: I don't quite follow this question... Is your objection that ``the set of monomial terms of $\beta_i$'' is too imprecise?

@@

However, I'm really baffled as to why you want to apply $\psi$ to the
$\beta_i$, since two lines later, you then apply the reverse correspondence
and want to work back in the semigroupring. Why don't you just stay in the
semigroup ring the whole time?
I think I'm misunderstanding this, but is $\omega_\mu$ just equal to
$\beta_i$ for some $i$? Also, why is $\omega_\mu$ missing a $d$ index
as its first index?
\\
Peter: $\beta_i$'s are polynomials in $f_1, \ldots f_4$.  By the monomial terms in $\beta_i$ I mean look at each monomial appearing in this polynomial and take the set of those.
}
can further compute that there exist $q_i, q_{i,j} \in \bq_{\geq 0}$ 
\todo{Aaron: Would it be reasonable to given these $q_{ij}$ a $\mu$
superscript or subscript to indicate dependence on $\mu$?}
and $\lambda \in
\Sigma$ so that the lattice point, $\omega_\mu = (c_1 ,\ldots c_n)$,
corresponding to $\mu
$ is of the form
\[
	\omega_\mu = \lambda + E_\mu
\]
where $\lambda$ is the same as in Equation \ref{eqn:sigma-in-terms-of-lambda/E} and $E_\mu$ is defined by

\begin{equation}\label{eqn:hirz-E_mu-translation}
E_\mu = \left(\sum_{\substack{i \in \{1, 2, 3, 4\} \cap \Te}}q_i e_i\right) +
\left(\sum_{\substack{i \in \{1, 2, 3, 4\} \cap \Tp \\ j\in \Tm}} q_{i,j}
e_{i,j}\right) + \left(\sum_{\substack{i \in \{1, 2, 3, 4\} \cap \Tm \\ j\in
\Tp}} q_{j,i} e_{i,j}\right).
\end{equation}

>>>>

Aaron: In general, I like the idea of having proofs self contained,
so that one should only refer to the proof statement, not the
body of the proof.
I'm not very strongly tied to this idea though, and I think it's fine. I think it's fine.
\\
Peter: I understand where your objection is coming from.  However, I don't see a good way to fix it without creating a very ugly statement of the lemma.
}
We note that $\Sigma$ has extremal rays given by the lattice points 
\begin{equation}\label{defn:e-i-proj}
	e_i = \left(\ell_i \bida_i, - \alpha_1 \ell_i \bida_i, \ldots
-\alpha_{i-1} \ell_i \bida_i, \ell_i \sum_{j\ne i} \alpha_j \bida_j,
-\alpha_{i+1} \ell_i \bida_i, \ldots, -\alpha_n, \ell_i \bida_i \right)
\end{equation}
for each $i\in \{1, \ldots n\}$.
Therefore, applying Proposition ~\ref{prop:cone-generation}, we see $R_D$ is generated in degrees less than
\[
	\sum_{i=1}^n \ell_i a_i.
\]
\end{proof}

@@

From this we can deduce that a set of extremal rays of $\Sigma$ are given as follows:
For $i \in \Te(D)$, define $e_i \in R_D$ in degree $d_i = \ell_i \gcd(a_i, b_i)$ by 
\begin{equation}\label{eqn:epsilon_i}
	e_i := d_i \epsilon_i %(-\alpha_1 d_i, \ldots, - \alpha_{i-1} d_i, \frac{d_i}{\bida_i + \bidb_i}\sum_{k \ne i}
	%\alpha_k (\bida_k + \bidb_k), - \alpha_{i+1} d_i, \ldots, - \alpha_n d_i)
\end{equation}

\noindent
For $i \in \Tp(D)$ and $j \in \Tm(D)$, with $i < j$, define $e_{i,j}\in R_D$ in degree $d_{i,j} = \ell_{i, j}(\bida_i \bidb_j - \bida_j \bidb_i)$ by
\begin{equation}\label{eqn:epsilon_i,j}
e_{i,j} := \frac{a_j\sum_{k\ne i,j} d_{i,j}\alpha_k b_k - b_j\sum_{k \ne i,j} d_k \alpha_k a_k}{a_ib_j - b_i a_j} \epsilon_i + \frac{a_i \sum_{k\ne i,j} d_{i,j} \alpha_k b_k - b_i\sum_{k\ne i,j} d_{i,j} \alpha_ka_k}{a_jb_i - b_j a_i} \epsilon_j
%	
%	(-\alpha_1 d_{i,j}, \ldots -\alpha_{i-1} d_{i,j}, \bidb_j \sum_{k \ne i,j} \alpha_k \bida_k - \bida_j \sum_{k \ne i, j}
%	\alpha_k \bidb_k, -\alpha_{i+1} d_{i,j}, \\
%	&\ldots -\alpha_{j-1} d_{i,j}, -\bidb_i \sum_{k \ne i, j} \alpha_k \bida_k + \bida_i \sum_{k \ne i, j}
%	\alpha_k \bidb_k, -\alpha_{j+1}d_{i,j}, \ldots, - \alpha_{n}d_{i,j}).
%	\end{split}
\end{equation}

@@

To find such a $\beta_i$,
\todo{Aaron: Shouldn't we say something above like, the kernel
	is bounded by the maximum of something over all $i$,
	so for the remainder of the proof, we fix an $i$.
	Also, I think it would help if things
	that depend on $i$, such as $E,\sigma,\lambda$
have an $i$ subscript.

Peter: Better (on the first comment? As for the second, that might be a little tricky, because we already have $E_\mu$, so we would have to do something like $E_{i,\mu}$ (which we could of course do...)}
we simply write $f_i$ as a 
polynomial in $f_1, \ldots f_4$. Furthermore, these
generate all relations, since they allow us to reduce any $\prod_{j =
1}^n f_j^{c_j}$ to the canonical form where $r_i = - \lfloor d \alpha_i
\rfloor$ whenever $i > 4$.  To bound the degree of generation of these relations, we bound the degree of generation of the ideal $(z_i - \beta_i) \cap \ker(\psi)$ for each $i$.

@@

Let $\omega\in \Omega'$ correspond to the monomial $y_i \prod_{j=1}^n
{y_j}^{c_j}$.  Then we can write $\omega$ as a sum of $r_j e_j$'s for $j\in
\Te$ and $r_{j,k}e_{j,k}$ for $j\in \Tp, k\in \Tm$.  
For convenience, let $e_j$ (when it exists) be in degree $d_j$ with $j^{th}$
component $-\alpha_j d_j + \kappa_j$ for some $\kappa_j \in \bq$ and let $e_{j,k}$ (when it exists) lie in
degrees $d_{j,k}$ with $j^{th}$ component $-\alpha_j d_{j,k} + \kappa_{j,k}'$ and $k^{th}$
component $-\alpha_j d_{j,k} + \kappa_{j,k}''$ for $\kappa_{j,k}', \kappa_{j,k}'' \in \bq$.
By a careful computation, we can find $\lambda \in \Sigma$ 
and $r_{i,j} \in \bq$ (in general such a choice may be very non-unique) so that we can write
\begin{equation}\label{eqn:sigma-in-terms-of-lambda/E}
	\sigma = \lambda + E
 \end{equation}
with $E$ of the form
\begin{equation}\label{eqn:hirz-E-translation}
	E = \begin{cases}
	\frac{1}{\kappa_i} e_i  	&  \mbox{ if }i\in \Te\\
	\sum_{j\in \Tm} r_{i,j} e_{i,j}: r_{i,j}\in \bq, \sum_{j\in \Tm} r_{i,j}\kappa_{i,j}' = 1 & \mbox{ if } i\in \Tp\\
	\sum_{j\in \Tp} r_{j,i} e_{j,i}: r_{j,i} \in \bq, \sum_{j\in \Tp} r_{j,i} \kappa_{i,j}'' = 1 	& \mbox{ if } i\in \Tm\\
	\end{cases}.
\end{equation}
Specifically, if we write out $\sigma = \sum_{j\in \Te} s_j e_j + \sum_{j\in \Tp, k \in \Tp} s_{j,k} e_{j,k}$, then any choice of $r_{j,k}$'s not greater than the respective $s_{j,k}$'s such that $r_{i,j} \kappa_{i,j}' =1$ (when $i\in \Tp$ but a similar statement holds for $i\in \Tm$) will suffice.  Such a choice of $r_{i,j}$'s uniquely determines the necessary $\lambda$.

From this, we can further deduce that for such an $E$,
\[
	\deg(E) \le \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm \end{cases}.
\]
Let $M$ be the set of monomials terms of $\beta_i$.  Then for $\mu = \prod_{j=1}^n
{f_i}^{c_i}\in M$, we
\todo{Aaron: What is $M$? \\
Peter: I don't quite follow this question... Is your objection that ``the set of monomial terms of $\beta_i$'' is too imprecise?

@@

However, I'm really baffled as to why you want to apply $\psi$ to the
$\beta_i$, since two lines later, you then apply the reverse correspondence
and want to work back in the semigroupring. Why don't you just stay in the
semigroup ring the whole time?
I think I'm misunderstanding this, but is $\omega_\mu$ just equal to
$\beta_i$ for some $i$? Also, why is $\omega_\mu$ missing a $d$ index
as its first index?
\\
Peter: $\beta_i$'s are polynomials in $f_1, \ldots f_4$.  By the monomial terms in $\beta_i$ I mean look at each monomial appearing in this polynomial and take the set of those.
}
can further compute that there exist $q_i, q_{i,j} \in \bq_{\geq 0}$ 
and $\lambda \in
\Sigma$ so that the lattice point, $\omega_\mu = (c_1 ,\ldots c_n)$,
corresponding to $\mu
$ is of the form
\[
	\omega_\mu = \lambda + E_\mu
\]
where $\lambda$ is the same as in Equation \ref{eqn:sigma-in-terms-of-lambda/E} and $E_\mu$ is defined by

\begin{equation}\label{eqn:hirz-E_mu-translation}
E_\mu = \left(\sum_{\substack{i \in \{1, 2, 3, 4\} \cap \Te}}q_i^\mu e_i\right) +
\left(\sum_{\substack{i \in \{1, 2, 3, 4\} \cap \Tp \\ j\in \Tm}} q_{i,j}^\mu
e_{i,j}\right) + \left(\sum_{\substack{i \in \{1, 2, 3, 4\} \cap \Tm \\ j\in
\Tp}} q_{j,i}^\mu e_{i,j}\right).
\end{equation}
